A pipeline to analyse time-course gene expression data

### 2019-05-10 scripts/manuscript.Rmd
Reactivating style
Putting back the whole manuscript

```{r schema, echo=FALSE, fig.width=8, fig.cap="Workflow for analyzing time-course datasets."}
knitr::include_graphics("images/fig1_schemas.png")
```

@@

## Overview

@@

## References

## Pre-processing and normalization

@@

```{r reading_targets, results="hide", fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
# Perform background correction using norm-exponential strategy and normalize
# across samples using quantile normalization.
RG = limma::backgroundCorrect(RG, method="normexp")
RG.q = limma::normalizeBetweenArrays(RG, method="quantile")
```

@@

```{r plotting_densities, results="hide", fig.width=8, fig.height=4}
# Plot the densities of the probes before and after normalization.
par(mfrow=c(1, 2))
limma::plotDensities(RG, col="black", legend=FALSE,
		     main="Before normalization")
limma::plotDensities(RG.q, col="black", legend=FALSE,
		     main="After normalization")
```

@@

## Exploratory analysis and quality control

@@

```{r echo=FALSE, fig.width=8, fig.height=4}
graphics::hist(variance_per_genes, breaks=100, col="black", xlab="Variance",
	       ylab="# probes")
graphics::abline(v=min_variance, col="#AB0000", lw=2)
```

@@

```{r pca, fig.width=8, fig.height=8}
# Reorder genes on condition, time, and replicate

@@

```{r pca_plots, fig.width=8, fig.height=8}
graphics::par(mfrow=c(2, 2))
graphics::plot(
    pca_data$x[, "PC2"], pca_data$x[, "PC1"],
    col=ann_colors$Group[meta$Group],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    xlab="PC2", ylab="PC1")
graphics::plot(
    pca_data$x[, "PC2"], pca_data$x[, "PC1"],
    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    xlab="PC2", ylab="PC1")

@@

```{r  correlation_plot, fig.width=6, fig.height=6}

@@

# Reorder genes on condition, time, and replicate
ord = order(
  meta$Group,
  meta$Timepoint,
  meta$Replicate)
variance_filtered_data = variance_filtered_data[, ord]
meta = meta[ord, ]

data_corr = stats::cor(variance_filtered_data, method="pearson")

# Copy the metadata before changing the timepoints into factors
data_corr_meta = meta
data_corr_meta$Timepoint = as.factor(meta$Timepoint)
NMF::aheatmap(
    data_corr,
    Colv=NA, Rowv=NA,
    annCol=data_corr_meta[, c("Group", "Timepoint")],
    annRow=data_corr_meta[, c("Group", "Timepoint")],
    annLegend=TRUE, 
    annColors=ann_colors,
    main="Correlation plot")
```

@@

# Differential expression analysis of time-course data

@@

## Weekly differential expression analysis

`moanin` provides a simple interface to perform a time by time differential
expression analysis. Under the hood, it simply calls `limma` on the set of
contrasts provided. Here, we show an example between the control mouse and the
mouse infected with the high dose of the influenza strain A/Vietnam/1203/04
(H5N1), but the function works with any form contrasts.

@@

```{r fig.width=8, fig.height=8, echo=FALSE}
par(mfrow=c(2, 2))

@@

timely_barplot = function(de_analysis, labels, qval_threshold=0.05, title=""){
    qval_colnames = colnames(de_analysis)[
	grepl("qval", colnames(de_analysis))]
    number_de_genes_per_time = colSums(de_analysis[, qval_colnames] < 0.05)
    names(number_de_genes_per_time) = labels
    barplot(number_de_genes_per_time, col="black", ylim=c(0, 20000),
	    xlab="Timepoint", ylab="Number of DE genes", main=title)

@@

## Time-course differential expression analysis between two groups

@@

```{r echo=FALSE, fig.width=6, fig.height=6}
num_de = colSums(qvalues < 0.05)
names(num_de) = c(
    "Kawasaki", "California", "Vietnam LD", "Vietname HD")
barplot(num_de, col="black", ylab="Number of DE genes")
```

@@

```{r echo=FALSE, fig.height=8, fig.width=8, message=FALSE}
par(mfrow=c(2, 2))
plot(log_fold_change_epicon[, "M-C"],
     log_fold_change_sum[, "M-C"], pch=20, main="EPICON vs sum",
     xlab="EPICON", ylab="sum")
plot(log_fold_change_epicon[, "M-C"],
     log_fold_change_max[, "M-C"], pch=20, main="EPICON vs max",
     xlab="EPICON", ylab="max")
plot(log_fold_change_max[, "M-C"],
     log_fold_change_min[, "M-C"], pch=20, main="max vs min",
     xlab="max", ylab="min")
plot(log_fold_change_max[, "M-C"],
     log_fold_change_sum[, "M-C"], pch=20, main="max vs sum",
     xlab="max", ylab="sum")
```

@@

```{r fig.height=8, fig.width=8}
pvalue = qvalues[, "M-K"]
lfc_epicon = log_fold_change_epicon[, "M-K"]
names(lfc_epicon) = row.names(log_fold_change_epicon)

@@

```{r fig.height=8, fig.width=8}
top_DE_genes_pval = names(sort(pvalue)[1:12])
moanin::plot_genes(data[top_DE_genes_pval, ], splines_model,
		   colors=ann_colors$Group, smooth=TRUE)
```

@@

```{r fig.height=8, fig.width=8}
top_DE_genes_lfc = names(
    sort(abs(lfc_epicon),
	 decreasing=TRUE)[1:12])
moanin::plot_genes(data[top_DE_genes_lfc, ], splines_model,
		   colors=ann_colors$Group, smooth=TRUE)

@@

# Clustering of time-course data

@@

```{r echo=FALSE, fig.height=6, fig.width=6, fig.caption="Confusion matrix between the two sets of labels"}
labels = unlist(labels[!is.na(labels)])
genes_scored = names(labels)

@@

## Looking at specific clusters in details.

@@

## How to choose the number of clusters.

@@

### Elbow method

@@

### Average silhouette method

@@

### Looking at the stability of the clustering

@@

#### The model explorer strategy

@@

#### Consensus clustering as a way to find $k$

@@

# Downstream analysis of clusters.

>>>>

```{r schema, echo=FALSE, fig.width=4, fig.cap="Workflow for analyzing time-course datasets."}

knitr::include_graphics("images/fig1_schemas.png")

@@

@@

## Quality control and normalization

### Pre-processing and normalization

@@

```{r reading_targets, results="hide", fig.width=4, fig.height=2, message=FALSE, warning=FALSE}
# Perform background correction using norm-exponential strategy and normalize
# across samples using quantile normalization.
RG = limma::backgroundCorrect(RG, method="normexp")
RG.q = limma::normalizeBetweenArrays(RG, method="quantile")
```

@@

```{r plotting_densities, results="hide", fig.width=6, fig.height=3}
# Plot the densities of the probes before and after normalization.
par(mfrow=c(1, 2), mar=c(2.0, 2.0, 2.0, 2.0))
limma::plotDensities(RG, col="black", legend=FALSE,
		     main="Before normalization", cex=0.8)
limma::plotDensities(RG.q, col="black", legend=FALSE,
		     main="After normalization", cex=0.8)
```

@@

### Exploratory analysis and quality control

@@

```{r echo=FALSE, fig.width=6, fig.height=3, fig.caption="Distribution of variance"}
graphics::hist(variance_per_genes, breaks=100, col="black", xlab="Variance",
	       ylab="# probes", main="")
graphics::abline(v=min_variance, col="#AB0000", lw=2)
```

@@

```{r pca, fig.width=6, fig.height=6}
# Reorder genes on condition, time, and replicate

@@

```{r pca_plots, fig.width=6, fig.height=6}
graphics::par(mfrow=c(2, 2), mar=c(2.5, 2.5, 2.5, 2.5))
graphics::plot(
    pca_data$x[, "PC2"], pca_data$x[, "PC1"],
    col=ann_colors$Group[meta$Group],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    xlab="PC2", ylab="PC1")
graphics::plot(
    pca_data$x[, "PC2"], pca_data$x[, "PC1"],
    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    xlab="PC2", ylab="PC1")

@@

```{r  correlation_plot, fig.width=6, fig.height=6, fig.caption="Cross-correlation of the samples"}

@@

# Reorder genes on condition, time, and replicate
ord = order(
  meta$Group,
  meta$Timepoint,
  meta$Replicate)

variance_filtered_data = variance_filtered_data[, ord]
data_corr = stats::cor(variance_filtered_data, method="pearson")

data_corr_meta = meta[ord, ]

# Copy the metadata before changing the timepoints into factors
data_corr_meta$Timepoint = as.factor(meta$Timepoint)
NMF::aheatmap(
    data_corr,
    Colv=NA, Rowv=NA,
    annCol=data_corr_meta[, c("Group", "Timepoint")],
    annRow=data_corr_meta[, c("Group", "Timepoint")],
    annLegend=TRUE, 
    annColors=ann_colors,
    main="Correlation plot")
```

@@

## Differential expression analysis of time-course data

@@

### Weekly differential expression analysis

`moanin` provides a simple interface to perform a time by time differential
expression analysis. Under the hood, it simply calls `limma` [@ritchie:limma] on the set of
contrasts provided. Here, we show an example between the control mouse and the
mouse infected with the high dose of the influenza strain A/Vietnam/1203/04
(H5N1), but the function works with any form contrasts.

**FIXME rename the function to something better**

@@

```{r fig.width=6, fig.height=6, echo=FALSE}
par(mfrow=c(2, 2))

@@

timely_barplot = function(de_analysis, labels, qval_threshold=0.05, title=""){
    qval_colnames = colnames(de_analysis)[
	grepl("qval", colnames(de_analysis))]
    number_de_genes_per_time = colSums(de_analysis[, qval_colnames] < 0.05)
    names(number_de_genes_per_time) = labels[1]
    barplot(number_de_genes_per_time, col="black", ylim=c(0, 20000),
	    xlab="Timepoint", ylab="Number of DE genes", main=title)

@@

### Time-course differential expression analysis between two groups

@@

```{r echo=FALSE, fig.width=3, fig.height=3, fig.caption="Distribution of DE genes per condition"}
num_de = colSums(qvalues < 0.05)
names(num_de) = c(
    "Kawasaki", "California", "Vietnam LD", "Vietname HD")
barplot(num_de, col="black", ylab="Number of DE genes")
```

@@

```{r echo=FALSE, fig.height=6, fig.width=6, message=FALSE}
par(mfrow=c(2, 2))
plot(log_fold_change_epicon[, "M-C"],
     log_fold_change_sum[, "M-C"], pch=20, main="EPICON vs sum",
     xlab="EPICON", ylab="sum")
plot(log_fold_change_epicon[, "M-C"],
     log_fold_change_max[, "M-C"], pch=20, main="EPICON vs max",
     xlab="EPICON", ylab="max")
plot(log_fold_change_max[, "M-C"],
     log_fold_change_min[, "M-C"], pch=20, main="max vs min",
     xlab="max", ylab="min")
plot(log_fold_change_max[, "M-C"],
     log_fold_change_sum[, "M-C"], pch=20, main="max vs sum",
     xlab="max", ylab="sum")
```

@@

```{r fig.height=4, fig.width=4}
pvalue = qvalues[, "M-K"]
lfc_epicon = log_fold_change_epicon[, "M-K"]
names(lfc_epicon) = row.names(log_fold_change_epicon)

@@

```{r fig.height=6, fig.width=6}
top_DE_genes_pval = names(sort(pvalue)[1:12])
moanin::plot_genes(data[top_DE_genes_pval, ], splines_model,
		   colors=ann_colors$Group, smooth=TRUE)
```

@@

```{r fig.height=6, fig.width=6}
top_DE_genes_lfc = names(
    sort(abs(lfc_epicon),
	 decreasing=TRUE)[1:12])
moanin::plot_genes(data[top_DE_genes_lfc, ], splines_model,
		   colors=ann_colors$Group, smooth=TRUE)

@@

## Clustering of time-course data

@@

```{r echo=FALSE, fig.height=6, fig.width=6, fig.caption="Confusion matrix between the two sets of labels"}
genes_scored = names(labels)

@@

### Looking at specific clusters in details.

@@

### How to choose the number of clusters.

@@

#### Elbow method

@@

#### Average silhouette method

@@

#### Looking at the stability of the clustering

@@

##### The model explorer strategy

@@

##### Consensus clustering as a way to find $k$

@@

## Downstream analysis of clusters.
